Preserving Reciprocal Consistency in Distributed Graph Databases

### 2020-04-07 paper/sec1-introduction.tex
camera-ready

When, for example, a BASE database \cite{Pritchett2008} is adapted with a graph processing layer, then violations of reciprocal consistency will occur if that adaptation provides no concurrency control for operations that span partitions in order to offer higher performance. This paper proposes a simple concurrency control protocol, called \tDelta protocol, that does not impede performance adversely. That is because the protocol is exclusively designed for one purpose only: reciprocal consistency in distributed edges.  Its design leverages the fact that the updating of end pointers of a given distributed edge immediately follow each other and the small interval between them is the window for possible conflicts between concurrent updates. The protocol ensures that any two updates on any given end pointer of any given edge are at least $\Delta$ time apart by aborting any write attempt that is within . It ensures reciprocal consistency when $\Delta$ is chosen to be larger than the length of the conflict window.

@@

\begin{figure}[ht!]
  \centering
  \includegraphics[width=\linewidth]{./figures/janusgraph-adj-list}
  \caption{Database records representing vertices, containing properties and a sequence of edges that comprise the adjacency list of a vertex \cite{janusgraph}.}
  \label{adj-list}
\end{figure}

>>>>

When, for example, a BASE database \cite{Pritchett2008} is adapted with a graph processing layer, then violations of reciprocal consistency will occur if that adaptation provides no concurrency control for operations that span partitions in order to offer higher performance. This paper proposes a simple concurrency control protocol, called \tDelta protocol, that does not impede performance adversely. That is because the protocol is exclusively designed for one purpose only: reciprocal consistency in distributed edges.  Its design leverages the fact that the updating of end pointers of a given distributed edge immediately follow each other and the small interval between them is the window for possible conflicts between concurrent updates.

@@

### 2020-04-07 paper/sec2-reciprocal-consistency.tex
camera-ready

A common approach to storing graphs (arising from JanusGraph \cite{janusgraph} and TitanDB \cite{TitanDB}) is given in Figure \ref{adj-list}.
A vertex is represented by a record that contains one or more vertex properties and a sequence of edge pointers (each shown simply as edge in Figure \ref{adj-list}) pointing to other vertices.
This sequence of edge pointers is referred to as an adjacency list.

Consider, for example, the statement that Tolkien \textit{wrote} The Hobbit.
It is expressed using vertices \emph{a} and \emph{b}, for Tolkien and The Hobbit respectively, and an edge \textit{wrote} running from \emph{a} (source) to \emph{b} (destination).

@@

Adjacency lists of both \emph{a} and \emph{b} record information about the edge and this information is reciprocal (or inverse) of each other: \emph{a}'s list will indicate `\emph{a} \emph{wrote} \emph{b}' while \emph{b}'s will have `\emph{b} \emph{written} by \emph{a}', Figure \ref{reciprocally-consistent-edge}. Thus, a query `list all titles by the author who wrote The Hobbit' can be answered starting at (destination vertex) \emph{b} and then traversing to (source vertex) \emph{a}, even though the edge is ``directed'' from \emph{a} to \emph{b} at model level abstraction. When the adjacency list entries for a given edge are mutually compatible like this, that edge is said to be \emph{reciprocally consistent}, a form of referential integrity. A query can read either the source or destination vertex and is able to reify the edge correctly, returning consistent results.

\input{figures/reciprocally-consistent-edge}

>>>>

Consider, for example, the statement that Tolkien \textit{wrote} The Hobbit.
It is expressed using vertices \emph{a} and \emph{b}, for Tolkien and The Hobbit respectively, and an edge \textit{wrote} running from \emph{a} (source) to \emph{b} (destination). The logical representation of the edge is given in Figure \ref{rc-edge}.

@@

A common internal representation of this graph (arising from JanusGraph \cite{janusgraph} and TitanDB \cite{TitanDB}) is given in Figure \ref{db-rep}.
A vertex is represented by a record that contains one or more vertex properties and a sequence of edge pointers pointing to other vertices.
This sequence of edge pointers is referred to as an adjacency list.

From Figure \ref{db-rep} it can be seen adjacency lists of both \emph{a} and \emph{b} record information about the edge and this information is reciprocal (or inverse) of each other: \emph{a}'s list will indicate `\emph{a} \emph{wrote} \emph{b}' while \emph{b}'s will have `\emph{b} \emph{written} by \emph{a}'.
Thus, a query `list all titles by the author who wrote The Hobbit' can be answered starting at (destination vertex) \emph{b} and then traversing to (source vertex) \emph{a}, even though the edge is ``directed'' from \emph{a} to \emph{b} at model level abstraction.
When the adjacency list entries for a given edge are mutually compatible like this, that edge is said to be \emph{reciprocally consistent}, a form of referential integrity.
A query can read either the source or destination vertex and is able to reify the edge correctly, returning consistent results.

\input{figures/rc}

### 2020-04-07 paper/sec5-delta.tex
camera-ready

The \tDelta Protocol has five rules:
\begin{enumerate}
\item A transaction's update on an end pointer of an edge is initially tentative which would become permanent only if that transaction is permitted to commit.
\item A tentative write is possible if the end pointer is either in a permanent state or the immediately preceding tentative update was done at least $\Delta$ time before (where time is measured as per local clock).
\item If a transaction successfully performs all its tentative updates, then it is permitted to commit; otherwise, it must abort.
\item A transaction commits, when all its tentative updates are made permanent, e.g., using an atomic commitment protocol.
\item Tentative updates of an aborting transactions are ignored. An ignored tentative update can make a new transaction abort for up to $\Delta$ time after it was created; it is harmless thereafter and can be garbage collected at any time.
\end{enumerate}

@@

Let us define $\delta$ as the bound estimate on the interval that may elapse between a transaction updating one end of a distributed edge at one server and updating the other end of the same edge at another server.

Let $\Delta$ be chosen such that $\Delta > \delta$. Now consider the interleaving in Figure \ref{fig:tx-ty} and let $t_x$ be the (global) time when $T_x$ starts at server $S_i$; similarly $t_y$ be the time $T_y$ starts at server $S_j$. Note $t_y < t_x$, i.e. $(t_x - t_y) > 0$.

Say, $T_y$ reaches $S_i$ at time $t_y + d$, where $d$ is the actual time elapsed between completing a tentative write at one end and starting at another end, let us assume that $d \leq \delta$. When $T_y$ arrives at $S_i$ it will find a tentative write done at time $t_x$. In this case, $t_y + d - t_x = d - (t_x-t_y) < d \leq \delta < \Delta$; so, $T_y$ will abort, preventing writes from interleaving and half-corrupting the edge. A similar arguments can be made for the scenario in with Figure \ref{fig:ty-tx}. For the interleaving in Figure \ref{fig:same-side}, if $t_y - t_x > \Delta$, then $T_y$ cannot overtake $T_x$ at $S_j$.

Assume that $d > \Delta$, i.e., the estimate $\delta$ does not hold at this moment. In Figure \ref{fig:tx-ty} interfering updates are avoided only if $t_y + d - t_x = d - (t_x-t_y) < d \leq \delta < \Delta$ otherwise, an edge becomes half-corrupted. This holds for the interleaving in \ref{fig:ty-tx}. Thus, in the extreme case $t_x = t_y$, reciprocal consistency is not guaranteed if $d > \Delta$ when the former exceeds its upper bound estimate $\delta$. For Figure \ref{fig:same-side}, if $d$ for $T_x$ is larger than $\Delta$, reciprocal consistency can be violated.

This protocol is strictly weaker than Read Uncommitted isolation, ensuring only writes to distributed edge are totally ordered, preserving reciprocal consistency. The benefit of this approach is reduced contention on frequently accessed distributed edges, as the time a transaction has exclusive access to a given distributed edge is decreased.

In summary, the \tDelta protocol attempts to prevent the occurrence of half-corrupted distributed edges by aborting transactions but if $\Delta$ is exceeded edge can become half-corrupted and process of corruption described in Section \ref{sec:db-corruption} can still occur. Two questions naturally arise regards the performance of the protocol. For different values of $\Delta$:
\begin{itemize}
\item Given distributed edges can still become half-corrupted if $\Delta$ is exceeded, by how much time does the protocol increase the time until operational corruption?
\item How many transactions are aborted as a result of the protocol?
\end{itemize}

>>>>

The \tDelta Protocol has five rules:
\begin{enumerate}
\item A transaction's write on an end pointer of an edge is initially tentative which would become permanent only if that transaction is permitted to commit.
\item A tentative write is possible if the end pointer is either in a permanent state or the immediately preceding tentative write was done at least $\Delta$ time before (where time is measured as per local clock).
\item If a transaction successfully performs all its tentative writes, then it is permitted to commit; otherwise, it must abort.
\item A transaction commits, when all its tentative writes are made permanent, e.g., using an atomic commitment protocol.
\item Tentative writes of an aborting transactions are ignored. An ignored tentative write can make a new transaction abort for up to $\Delta$ time after it was created; it is harmless thereafter and can be garbage collected at any time.
\end{enumerate}

@@

Let us define $\delta$ as the bound estimate on the interval that may elapse between a transaction writing one end of a distributed edge at one server and writing the other end of the same edge at another server.

Let $\Delta$ be chosen such that $\Delta > \delta$.
Now consider the interleaving in Figure \ref{fig:b} and let $t_x$ be the (global) time when $T_x$ starts at server $S_i$; similarly $t_y$ be the time $T_y$ starts at server $S_j$.
Note $t_y < t_x$, i.e. $(t_x - t_y) > 0$.

Say, $T_y$ reaches $S_i$ at time $t_y + d$, where $d$ is the actual time elapsed between completing a tentative write at one end and starting at another end, let us assume that $d \leq \delta$.
When $T_y$ arrives at $S_i$ it will find a tentative write done at time $t_x$.
In this case, $t_y + d - t_x = d - (t_x-t_y) < d \leq \delta < \Delta$; so, $T_y$ will abort, preventing writes from interleaving and half-corrupting the edge.
A similar arguments can be made for the scenario in with Figure \ref{fig:b}, where $t_y > t_x$ and $T_x$ will abort because it will find out, on reaching server $S_j$ for the final tentative write, that $t_x +d - t_y = d - (t_y-t_x) < d <= \delta < \Delta$.
For the interleaving in Figure \ref{fig:same-side}, if $t_y - t_x > \Delta$, then $T_y$ cannot overtake $T_x$ at $S_j$.

Assume that $d > \delta$, i.e., the estimate $\delta$ does not hold at this moment.
In Figure \ref{fig:b} interfering writes are avoided only if $t_y + d - t_x = d - (t_x-t_y) < d < \Delta$ otherwise, reciprocal inconsistency occurs.
In Figure \ref{fig:a} interfering writes are avoided only if $t_x + d - t_y = d - (t_y-t_x) < d < \Delta$ otherwise, reciprocal inconsistency occurs.
Thus, in the extreme case $t_x = t_y$ and reciprocal consistency is not guaranteed if $\Delta < d$ when the latter exceeds its upper bound estimate $\delta$.
For Figure \ref{fig:c}, interfering writes are avoided only if $(t_y + d) - (t_x + d) = (t_y- t_x) < d < \Delta$ other reciprocal inconsistency occurs.

% This protocol is strictly weaker than Read Uncommitted isolation, ensuring only writes to distributed edge are totally ordered, preserving reciprocal consistency. The benefit of this approach is reduced contention on frequently accessed distributed edges, as the time a transaction has exclusive access to a given distributed edge is decreased.

% In summary, the \tDelta protocol attempts to prevent the occurrence of half-corrupted distributed edges by aborting transactions but if $\Delta$ is exceeded edge can become half-corrupted and process of corruption described in Section \ref{sec:db-corruption} can still occur. Two questions naturally arise regards the performance of the protocol. For different values of $\Delta$:
% \begin{itemize}
% \item Given distributed edges can still become half-corrupted if $\Delta$ is exceeded, by how much time does the protocol increase the time until operational corruption?
% \item How many transactions are aborted as a result of the protocol?
% \end{itemize}

In summary, the \tDelta protocol eliminates interleaving of transactions during edge writes, so long as $\Delta$ remains larger than the interval $d$ that elapses between a transaction completing its write at one end of a distributed edge and starting at the other end.
Since the exact value of $d$ taken by a transaction cannot be precisely known in advance, its bound $\delta$ is estimated with the best effort and we choose $\Delta > \delta$.

The larger the value of $\Delta$ used, the more likely is that $\Delta > d$ holds and half-corruption and thereby operational corruption are averted; also, the more likely is that one or more of concurrent transactions will find their tentative write within $\Delta$ time of each other and choose to abort.
Our performance evaluation of the protocol will therefore involve measuring the folowing two metrics for various values of $\Delta$:
\begin{itemize}
\item Time taken for 10\% of a large database to be corrupted due to the introduction of half-corruption, and
\item Number of transactions being aborted per second.
\end{itemize}
e let $d$ be an exponentially distributed random variable with mean $\mu$, $D\sim \exp(\mu)$. Therefore, the probability of $d$ exceeding $\Delta$ is given by $P(D > \Delta) = e^{(-\mu \Delta)}$.

### 2020-04-07 paper/sec6-modeling.tex
camera-ready

The system processes transactions that arrive in a Poisson stream with rate $\lambda$ per second. To simplify the model, each transaction contains a random number of read operations, $K$, followed by a single write. To model a scale-free graph, edges in the database are divided into $T$ types, popular edges types have higher access probabilities but are a smaller proportion of the total number of edges, $N$. For each type, a fraction $f$ are distributed edges and the remainder are local edges. The network delay between servers was assumed to be exponentially distributed with rate $\rho$. At any moment in time an edge can be in one of four states:
\begin{enumerate}
\item Local and clean.
\item Distributed and clean.
\item Half-corrupted distributed edge arising from a dirty write.
\item Semantically corrupted.
\end{enumerate}
% The valid state transitions are given in Figure \ref{state-transitions}. Note, only distributed edges can be in state 2, but any edge, including local ones, can be in state 3.

@@

The \tDelta protocol impacts the rate of corruption by reducing the probability a transaction corrupts a distributed edge, $q^{new}_i$. Of interest, therefore, is: how large or small is the value of $U_\gamma$ for a given value of $\gamma$ under the \tDelta protocol? Moreover, to quantify the number of aborts, a second simulation which focused specifically on the subset of frequently accessed distributed edges was performed. The answers to these questions depends on several parameters characterizing the following systemic aspects:
\begin{itemize}
\item \emph{Database Size}. Size is expressed by the total number of edges $N$, and the fraction $f$ of distributed edges.
\item \emph{Workload}. Measured as transactions per second (TPS). Significant for measuring $U_{\gamma}$ are: the fraction of this load that writes after reads and the number of reads that precede a write.
\item \emph{Distributed Write Delays and Choosing $\Delta$}. The smaller the delays the less likely the bound $\Delta$ is violated. Conversely, smaller $\Delta$ is the more likely the bound $\Delta$ is violated.
\end{itemize}

>>>>

The system processes transactions that arrive in a Poisson stream with rate $\lambda$ per second.
To simplify the model, each transaction contains a random number of read operations, $K$, followed by a single write.
To model a scale-free graph, edges in the database are divided into $T$ types, popular edges types have higher access probabilities but are a smaller proportion of the total number of edges, $N$.
For each type, a fraction $f$ are distributed edges and the remainder are local edges.
The delay $d$ between a transaction completing a tentative write at one end and starting at another end is exponential distributed with a mean of $\mu$.
At any moment in time an edge can be in one of four states:
\begin{enumerate}
\item Local and clean.
\item Distributed and clean.
\item Half-corrupted distributed edge arising from a dirty write.
\item Semantically corrupted.
\end{enumerate}
% The valid state transitions are given in Figure \ref{state-transitions}. Note, only distributed edges can be in state 2, but any edge, including local ones, can be in state 3.

@@

The \tDelta protocol impacts the rate of corruption by reducing the probability a given transaction half-corrupts a distributed edge, $q^{new}_i$, see Appendix \ref{sec:deriv-confl-prob} for the derivation. Of interest, therefore, is: how large or small is the value of $U_\gamma$ for a given value of $\gamma$ under the \tDelta protocol? Moreover, to quantify the number of aborts, a second simulation which focused specifically on the subset of frequently accessed distributed edges was performed. The answers to these questions depends on several parameters characterizing the following systemic aspects:
\begin{itemize}
\item \emph{Database Size}. Size is expressed by the total number of edges $N$, and the fraction $f$ of distributed edges.
\item \emph{Workload}. Measured as transactions per second (TPS). Significant for measuring $U_{\gamma}$ are: the fraction of this load that writes after reads and the number of reads that precede a write.
\item \emph{Distributed Write Delays and Choosing $\Delta$}. The smaller the delays the less likely the bound $\Delta$ is violated. Conversely, smaller $\Delta$ is the more likely the bound $\Delta$ is violated.
\end{itemize}

### 2020-04-07 paper/sec7-evaluation.tex
camera-ready

The graph analyzed consisted of seven edge types, $N_1=10^4, N_2=10^5, N_3=10^6, N_4=10^7,  N_5=10^8, N_6=10^9, N_7=10^{10}$, totaling 11 billion edges, with access probabilities $p_1 =0.5, p_2 =0.25, p_3=0.13, p_4=0.06, p_5=0.03, p_6=0.02$ and $p_7 =0.01$; a graph of this size would have approximately 1 billion vertices. The number of read operations per query is geometrically distributed starting at $2$, with an average of $15$, before a write. In all edge types, a fraction $0.3$ are distributed, the remainder are local; in proportion with a good graph partitioning algorithm. The network delay between servers is exponential distributed with a rate of $200ms$. The database is initial clean and considered to be corrupted when $10$\% ($\gamma = 0.1$) of all edges are corrupted. The time taken until operational corruption $U$, is measured in days. $U$ considered for a range of transaction arrival rates, $\lambda = (1000, ..., 10000)$; a typical graph workload comprises of 90\% read-only transactions and 10\% read-write transactions \cite{Angles2020}, hence the chosen range reflects a total workload of $\lambda = (10000, ..., 100000)$. The following $\Delta$ values were considered $\Delta = 50, 75, 100ms$.

The results for measuring the impact of $\Delta$ on the time until operational corruption are given in Figure \ref{time-to-corruption-results}. Under no isolation, $U$ ranges between 500-50 days. For $50ms$ this increases to 74-1 years. For $\Delta = 75, 100ms$ the time to corruption vastly exceeds lifetime of any system making data corruption resulting from half-corrupted edges of little practical concern.

\begin{figure}[h]
  \centering
  \includegraphics[width=\linewidth]{./figures/delta}
  \caption{Time, $\log(U)$, until operational corruption ($\gamma = 0.1$) for $\Delta = 50, 75, 100ms$.}
  \Description{Graph representation.}
  \label{time-to-corruption-results}
\end{figure}

The abort rates for $\Delta = 50, 75, 100ms$ are given in Figure \ref{aborts-results} for the most popular edge type, $N=10^4$. The simulation was ran for 10 seconds for a range of transaction arrival rates, $\lambda = (1000, ..., 10000)$. For $\Delta = 50$ the abort rate varies between $1-5\%$, this increases to between $1-7\%$ and $1-9\%$ for $\Delta = 75$ and $\Delta = 100$ respectively.

\begin{figure}[h]
  \centering
  % \includegraphics[width=\linewidth]{./figures/aborts_50_75_100}
 \includegraphics[width=\linewidth]{./figures/aborts}
  \caption{Aborts per seconds for $\Delta = 50, 75, 100ms$.}
  \Description{Graph representation.}
  \label{aborts-results}
\end{figure}

>>>>

The graph analyzed consisted of seven edge types, $N_1=10^4, N_2=10^5, N_3=10^6, N_4=10^7,  N_5=10^8, N_6=10^9, N_7=10^{10}$, totaling 11 billion edges, with access probabilities $p_1 =0.5, p_2 =0.25, p_3=0.13, p_4=0.06, p_5=0.03, p_6=0.02$ and $p_7 =0.01$; a graph of this size would have approximately 1 billion vertices.
The number of read operations per query is geometrically distributed starting at $2$, with an average of $15$, before a write.
In all edge types, a fraction $0.3$ are distributed, the remainder are local; in proportion with a good graph partitioning algorithm.
The delay $d$ between a transaction completing a tentative write at one end and starting at another end is exponential distributed with a mean of $5ms$.
The database is initial clean and considered to be corrupted when $10$\% ($\gamma = 0.1$) of all edges are corrupted.
The time taken until operational corruption $U$, is measured in days.
$U$ considered for a range of transaction arrival rates, $\lambda = (1000, ..., 10000)$; a typical graph workload comprises of 90\% read-only transactions and 10\% read-write transactions \cite{Angles2020}, hence the chosen range reflects a total workload of $\lambda = (10000, ..., 100000)$.
The following $\Delta$ values were considered $\Delta = 50, 75, 100ms$.
For each $\Delta$ the probability that $d$ exceeds $\Delta$ is $P(D > \Delta) = 4.5 \time 10^{-5}, 3.1 \times 10^{-7}, 2.1 \times 10^{-9}$ respectively.

The results for measuring the impact of $\Delta$ on the time until operational corruption are given in Figure \ref{time-to-corruption-results}.
Under no isolation, $U$ ranges between 50-500 days.
For $50ms$ this increases to 1-75 years. For $\Delta = 75, 100ms$ the time to corruption vastly exceeds lifetime of any system making data corruption resulting from half-corrupted edges of little practical concern.

\begin{figure}[h]
  \centering
  \includegraphics[width=\linewidth]{./figures/delta}
  \caption{Time until operational corruption.}
  \label{time-to-corruption-results}
\end{figure}

The number of aborts per second for $\Delta = 50, 75, 100ms$ are given in Figure \ref{aborts-results} for the most popular edge type, $N=10^4$. The simulation was ran for 10 seconds for a range of transaction arrival rates, $\lambda = (1000, ..., 10000)$. For $\Delta = 50$ the abort rate varies between $1-5\%$, this increases to between $1-7\%$ and $1-9\%$ for $\Delta = 75$ and $\Delta = 100$ respectively.

\begin{figure}[h]
  \centering
  \includegraphics[width=\linewidth]{./figures/aborts}
  \caption{Number of aborts per second.}
  \label{aborts-results}
\end{figure}
